PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-12 IEEEtran/paper-annotated.tex
removed annex
Update

\TG{I'm still unsure what to do with this part, it looks out of place.}
\add{PyTracer follows a three steps approach:}
\begin{enumerate}
    \item \add{Automatic instrumentation of Python programs to save traces containing input and output values of all functions.}
    \item \add{Traces aggregation to produce a statistics summary.}
    \item \add{Interactive visualization to navigate through code execution.}
\end{enumerate}

@@

\add{
    The trace-based approach used by PyTracer permits to limit the 
    overhead at runtime since the statistics computation is reported to the postprocess analysis.
    However, it limits the aggregation to deterministic executions,
    although control-flow analysis can help to merge divergent execution paths.
}

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}.
First, \pytracer uses \textit{monkey patching} to replace the functions called
by the application with a wrapper that saves their arguments and returned values
in a trace file.
Second, it executes the application multiple times with noise injection, and
aggregates the resulting trace files to compute summary statistics about each
traced variable. Finally, a Plotly dashboard provides interactive visualizations
highlighting unstable code sections.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
simulates roundoff errors by applying function $inexact$ only to the
outputs returned by functions. The Precision Bounding (PB) mode can
identify other types of instabilities, in particular catastrophic
cancellations, however, it is more difficult to apply to large code bases
as it often leads to execution errors that require human intervention.}

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high numerical precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.}

@@

@@

\appendix

\include{annex}

>>>>

\TG{I'm still unsure what to do with this part, it looks out of place.}
% \add{PyTracer follows a three steps approach:}
% \begin{enumerate}
%     \item \add{Automatic instrumentation of Python programs to save traces containing input and output values of all functions.}
%     \item \add{Traces aggregation to produce a statistics summary.}
%     \item \add{Interactive visualization to navigate through code execution.}
% \end{enumerate}

\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}.
First, \pytracer uses \textit{monkey patching} to replace the functions called
by the application with a wrapper that saves their arguments and returned values
in a trace file.
Second, it executes the application multiple times with noise injection, and
aggregates the resulting trace files to compute summary statistics about each
traced variable. Finally, a Plotly dashboard provides interactive visualizations
highlighting unstable code sections.

@@

\add{
    The trace-based approach used by PyTracer permits to limit the 
    overhead at runtime since the statistics computation is reported to the postprocess analysis.
    However, it limits the aggregation to deterministic executions,
    although control-flow analysis can help to merge divergent execution paths.
    Hence, PyTracer is restricted to single-threaded applications even if its usage
    is limited in Python due to the Global Interperter Lock (GIL) mechanism
    that allows only one thread to hold the control of the Python interpreter.
}

% \subsection{Pytracer workflow}

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
    simulates roundoff errors by applying function $inexact$ only to the
    output returned by a function. The Precision Bounding (PB) mode can
    identify other types of instabilities, in particular catastrophic
    cancellations, however, it is more difficult to apply to large code bases
    as it often leads to execution errors that require human intervention.}

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high numerical precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that uses
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low precision (below 0).
    Following Equation}~\ref{eq:sig-digits}, \add{a number of significant digits below 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

@@

\subsection{\pytracer instrumentation limitations}

\add{
    The limitations of the PyTracer's instrumentation have little impact in practice
    since we have encountered few classes having non-writeable attributes in practice.
    Moreover, those classes are generally used in higher-level classes written in
    pure Python. For instance, the SGD class that implements the Stochastic Gradient
    Descent in scikit-learn takes a loss function as a parameter. Loss functions are
    implemented in Cython in scikit-learn and Pytracer is not able to instrument
    them. However, Pytracer can dump the instance of LossFunction if they are used
    by the SGD class as arguments, which is a frequently observed pattern in
    Scikit-learn for instance. So one still has an estimation of variability for the
    attributes of the class. Finally, one can use Veritracer to trace the methods of
    those Cython classes for an in-depth analysis since it is the C language
    underhood. We have encountered few iterators and none of them involved floating
    point values so we were able to safely ignore them. 
}

@@

% \appendix
% \include{annex}
